Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-02-05 WSSSPE3/report/wg_appendix_eng_design.tex
Section 5.5 and Appendix G: tiny changes.

Birgit Penzenstadler\footnote{email:
\href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}}
and Colin C. Venters\footnote{email:
\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points
of contact for this working group, and be responsible for ensuring timely
progress of the planned actions.

@@

\begin{itemize}
\item Birgit Penzenstadler -- California State University, CA, USA
\item Colin C. Venters -- University of Huddersfield, Huddersfield, UK
\item Matthias Bussonnier -- UC Berkeley, CA, USA
\item Jeff McWhirter -- Geode Systems 
\item Patrick Nichols -- National Center for Atmospheric Research, CO, USA
\item Ilian Todorov -- Science \& Technology Facilities Council, UK
\item Ian Taylor -- Cardiff University, UK
\item Alexander Vyushkov -- University of Notre Dame, IN, USA
\end{itemize}

@@

The group included members from different backgrounds, including quantum
chemistry, epidemiology, computer science, software engineering, and microscopy.
Each participant was invited to give their perspective on the topic area and
what they thought were the crucial points for discussion. There was a general
consensus that there was a need for relating principles to practice for the
computational science and engineering community. Furthermore, various members of
the group expressed their interest in tools and best practices for facilitating
the maintenance and evolution of scientific software systems. It was agreed to
identify principles from software engineering and from sustainability design
and, based on those lists, discuss what each of those would mean applied to
specific example systems from the expert domains of some of the group members.
The group identified a number of software engineering principles drawn from the
Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles included: abstraction; coupling and cohesion;
decomposition and modularization; encapsulation and information hiding;
separation of interface and implementation; sufficiency completeness \&
primitiveness; and separation of concerns. Similarly, user interface design
principles included: learnability; user familiarity; consistency; minimal
surprise; recoverability; user guidance; and user diversity. The sustainability
design principles were drawn from the Karlskrona Manifesto on Sustainability
Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic;
multidimensional; interdisciplinary; transcends the system's purpose; applies to
both a system and its wider contexts; requires action on multiple levels;
requires multiple timescales; changing design to take into account long-term
effects doesn't automatically imply sacrifices; system visibility is a
precondition for and enabler of sustainability design. A number of sustainable
software engineering principles proposed by Tate~\cite{tate:2005} were also
considered including: continual refinement of product and project practices; a
working product at all times; continual emphasis on design; and value defect
prevention over defect detection.

@@

\subsection{Key Next Steps}
In order to achieve (1) the systematic analysis a number of example systems from
different scientific domains with regards to the identified principles, (2) the
identification of the commonalities and gaps in applying principles to different
scientific systems, and (3) the proposal of a set of guidelines on the
principles, the following next steps were discussed:

\subsection{Plan for Future Organization}
The following plan for future organization was discussed:
\begin{itemize}
\item Identify suitable undergraduate or post-graduate students.
\item Design and pilot study.
\item Organizing coordinating online calls via Google Hangout.
\end{itemize}

\subsection{What Else is Needed?}
\begin{itemize}
\item Ethics committee review panel approval required for data collection.
\end{itemize}

\subsection{Key Milestones and Responsible Parties}
The following key milestones were discussed as a roadmap for the set of
guidelines on software engineering principles:
\begin{itemize}
\item Oct/Nov 2015: Study design and interview guideline
\item Jan/Feb 2016: Interviews conducted and transcribed
\item Mar/Apr 2016: Analysis complete
\item May 2016: Report written
\end{itemize}

\subsection{Description of Funding Needed}
Specific funding was not discussed in this working group. However, this is a
open topic that can be discussed in relation to emerging funding calls from
National agencies or grant proposal initiatives.

>>>>

Birgit Penzenstadler\footnote{email:
\href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}}
and Colin C.\ Venters\footnote{email:
\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points
of contact for this working group, and be responsible for ensuring timely
progress of the planned actions.

@@

\begin{itemize}
\item Birgit Penzenstadler -- California State University, CA, USA
\item Colin C.\ Venters -- University of Huddersfield, Huddersfield, UK
\item Matthias Bussonnier -- UC Berkeley, CA, USA
\item Jeff McWhirter -- Geode Systems 
\item Patrick Nichols -- National Center for Atmospheric Research, CO, USA
\item Ilian Todorov -- Science \& Technology Facilities Council, UK
\item Ian Taylor -- Cardiff University, UK
\item Alexander Vyushkov -- University of Notre Dame, IN, USA
\end{itemize}

@@

This group is comprised  of members from different backgrounds, including quantum
chemistry, epidemiology, microscopy, computer science, and software engineering.
Each participant was invited to give their perspective on the topic area and
what they thought were the crucial points for discussion. There was a general
consensus that there was a need for relating principles to practice for the
computational science and engineering community. Furthermore, various members of
the group expressed their interest in tools and best practices for facilitating
the maintenance and evolution of scientific software systems. It was agreed to
identify principles from software engineering and from sustainability design
and, based on those lists, discuss what each of those would mean applied to
specific example systems from the expert domains of some of the group members.
The group identified a number of software engineering principles drawn from the
Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.

Software design principles include abstraction; coupling and cohesion;
decomposition and modularization; encapsulation and information hiding;
separation of interface and implementation; sufficiency completeness and
primitiveness; and separation of concerns. Similarly, user interface design
principles include learnability; user familiarity; consistency; minimal
surprise; recoverability; user guidance; and user diversity. The sustainability
design principles were drawn from the Karlskrona Manifesto on Sustainability
Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic,
multidimensional, and interdisciplinary; transcends the system's purpose; applies to
both a system and its wider contexts; requires action on multiple levels;
requires multiple timescales; changing design to take into account long-term
effects does not automatically imply sacrifices; system visibility is a
precondition for and enabler of sustainability design. A number of sustainable
software engineering principles proposed by Tate~\cite{tate:2005} were also
considered including: continual refinement of product and project practices; a
working product at all times; continual emphasis on design; and value defect
prevention over defect detection.

@@

\subsection{Key Next Steps}

In order to achieve the following three goals: (1) a systematic analysis of a number of example systems from
different scientific domains with regards to the identified principles, (2) the
identification of the commonalities and gaps in applying the principles to different
scientific systems, and (3) a proposal of a set of guidelines on the
principles, the next steps were discussed in the following subsections.

\subsection{Plan for Future Organization}

The following plan for future organization was discussed:
\begin{itemize}
\item Identify suitable undergraduate or post-graduate students.
\item Design and pilot study.
\item Organize coordinating online calls via Google Hangout.
\end{itemize}

\subsection{What Else is Needed?}

\begin{itemize}
\item Ethics committee review panel approval required for data collection.
\end{itemize}

\subsection{Key Milestones and Responsible Parties}

The following key milestones were discussed as a roadmap for the set of
guidelines on software engineering principles:
\begin{itemize}
\item Oct/Nov 2015: Study design and interview guideline
\item Jan/Feb 2016: Interviews conducted and transcribed
\item Mar/Apr 2016: Analysis complete
\item May 2016: Report written
\end{itemize}

\subsection{Description of Funding Needed}

Specific funding was not discussed in this working group. However, this is a
open topic that can be discussed in relation to emerging funding calls from
National agencies or grant proposal initiatives.

### 2016-02-05 WSSSPE3/report/wg_main_eng_design.tex
Section 5.5 and Appendix G: tiny changes.

Software engineering principles form the basis of methods, techniques,
methodologies and tools~\cite{Vliet:2008}. However, there is often a mismatch
between software engineering theory and practice particularly in the fields of
computational science and engineering, which can lead to the development of
unsustainable software~\cite{Merali:2010,hettrick:2014}. Understanding and
applying software engineering principles is essential in order to create and
maintain sustainable software~\cite{Becker:2016}.

@@

Software engineering principles form the basis of methods, techniques,
methodologies and tools. This group, which included members from different
backgrounds, including quantum chemistry, epidemiology, computer science,
software engineering, and microscopy, discussed the principles of software
engineering design for sustainable software (starting with principles from the
Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015},
Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
(SWEBOK)~\cite{swebokv3}) and their application in various domains including
quantum chemistry and epidemiology. The group examined the principles and took a
retrospective analysis of what the developers did in practice against how the
principles could have made a difference, and asked, what do the principles mean
for computational scientific and engineering software, and how do the principles
relate to non-functional requirements? It appeared that the sustainable software
engineering principles should be mapped to two core quality attributes that
underpin technically sustainable software: extensibility, the software's ability
to be extended and the level of effort required to implement the extension; and
maintainability: the effort required to locate and fix an error in operational
software.

@@

The next steps in this endeavor are to (1) Systematically analyze a number of
example systems from different scientific domains with regards to the identified
principles, to (2) Identify the commonalities and gaps in applying those
principles to different scientific systems, and to (3) Propose a set of
guidelines on the principles and how they exemplary apply to scientific software
system. Preliminary work will be carried out through undergraduate or
post-graduate student projects.

>>>>

Principles for software engineering form the basis of methods, techniques,
methodologies and tools~\cite{Vliet:2008}. However, there is often a mismatch
between software engineering theory and practice particularly in the fields of
computational science and engineering, which can lead to the development of
unsustainable software~\cite{Merali:2010,hettrick:2014}. Understanding and
applying software engineering principles is essential in order to create and
maintain sustainable software~\cite{Becker:2016}.

@@

Software engineering principles form the foundation of methods, techniques,
methodologies, and tools.  Consisting of members from different
backgrounds, including quantum chemistry, epidemiology, computer science,
software engineering, and microscopy, this group discussed the principles of software
engineering design for sustainable software (starting with principles from the
Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015},
Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
(SWEBOK)~\cite{swebokv3}) and their applications in various domains including
quantum chemistry and epidemiology. The group examined the principles and took a
retrospective analysis of what the developers did in practice against how the
principles could have made a difference, and asked, what do the principles mean
for computational scientific and engineering software, and how do the principles
relate to non-functional requirements? It appeared that the sustainable software
engineering principles should be mapped to two core quality attributes that
underpin technically sustainable software: \emph{extensibility}, the software's ability
to be extended and the effort level required to implement the extension; and
\emph{maintainability}: the effort required to locate and fix an error in operational
software.

@@

The next steps in this endeavor are to (1) Systematically analyze a number of
example systems from different scientific domains with regards to the identified
principles, to (2) Identify the commonalities and gaps in applying those
principles to different scientific systems, and to (3) Propose a set of
guidelines on the principles and examine how they exemplarily apply to scientific software
systems. Preliminary work will be carried out through undergraduate or
post-graduate student projects.
